Researchers collaborate with same-gendered colleagues more often than expected across the life sciences

### 2018-10-29 manuscript/Response_letter.Rmd

3.- The article reports a statistically significant linear relationship between standardized journal impact factor and $\alpha'$. However, there is a lot of noise and the slope of the corresponding regression is approx. 0.012. This looks like a rather weak relationship. Does it really worth being stressed so much?

**We don't think that we stress this result much. The impact factor result is only discussed in two very cautious sentences in the Discussion (cautious words shown in bold):**

We also **found a little evidence** that gender homophily is detrimental to research quality, in that high-impact journals **tended to have** weaker homophily. **Assuming** that papers published in high-impact journals are of higher average quality **(which is contentious; [67])**, our results **provide non-experimental support for the hypothesis** that mixed-gender teams produce better research than single-gender teams [42–48].

**Please let us know if any of that still seems excessive upon a second reading. Also, a regression slope of -0.012 is actually quite a lot, for this dataset. The mean $\alpha'$ across journals is about 0.07, and so this slope means that increasing the journal impact factor by one standard deviation above the mean decreases $\alpha'$ from about 0.070 to 0.058; in proportional terms, this is 0.058 / 0.07 = 83% of the mean. The $R^2$ is 4%, meaning that if we know the journal impact factor, we can explain about 4% of the variance in $\alpha'$, which is not huge, but is perhaps bigger than one would expect given the large number of unmeasured covariates that also affect the two variables being correlated here. At any rate, we are very upfront in saying the relationship is noisy, we give all the statistics and confidence intervals needed to interpret the result, and we are careful to remind readers that correlation is not causality.**

@@

4.- What does ‘significant heterogeneity’ mean in line 141? I guess they are located above the 95% CI, but please specify.

**Line 141 (in the original draft) actually says ‘significant heterophily’, where heterophily is our term for an excess of opposite-sex coauthor pairs relative to the randome expection (defined in the Introduction). We define 'significant' heterophily as the case where our measure of assortment by gender, $\alpha$, is negative and has 95% confidence limits that lie entirely below zero. We explain this definition of statistical significance in the Methods (page 15).**

@@

- line 89 Author order differs across disciplines, e.g. it is alphabetical in maths. Please specify that you focus on life-sciences journals.

**We already specify that PubMed focuses on life sciences journals, as the reviewer notes elsewhere. Also, the presence of some alphabetically-ordered papers in the set would not alter the fact that the average seniority of last authors is greater than that of first authors, in most/all of the journals we studied.**

#### Methods
- line 265 Please specify how gender was assessed (even though this may already be in [5]). In particular, was it done automatically or manually? There may be issues with automatic gender assignment for authors of different geographic origins.

**We prefer to leave the full details to our previous paper, which has extensive descriptions, two different quality controls, and the full R code that was used to programmatically identify gender using authors' first names. Given our sample size (36 million authors), it hopefully goes without saying that automatic methods were used, but we now point this out explicitly in the Methods. Our earlier paper has extensive discussion of the geographic origin issue, and success rates for each of >100 different countries (almost all of which are over 70%, with China and Korea being significant exceptions among the major research-producing nations).**

@@

- line 352 log (no capital L)

**Done**

- line 362 Why would senior authors be the main drivers? It could as well be junior authors who chose their advisor...

**Indeed, the purpose of the first author vs last author analysis was to see if we can get some clues as to whether homophily is driven mostly by senior or junior scientists: we outlined both possibilities in this section. I think you were misled because the two sentences beginning on line 362 use the word "Assume" rhetorically; the sentences basically say, "Assuming senior researchers drive the results, we predict X. This is because of Y." We do not start from the assumption that senior researchers are responsible for homophily. We also began that paragraph by pointing out that the 'senior researchers drive homophily' hypothesis is already predominant in the literature (though to our knowledge it has not been effectively tested before).**

@@

-- Right-hand side figure: what do the dots correspond to? a single journal?

**Yes -- we now state this in the figure legend.**

  - Do you identify individual authors/ Control for the fact that some are more prolific than others?

**No, but this would not affect the veracity of our findings. Let's imagine that 90% of all papers in the world had one very prolific individual in their author lists. Let's assume this author is a man, though the argument would be the same if they were a woman. The existence of this prolific man would increase the proportion of male co-authors for both sexes, since most people have co-published with him. However, it would not affect our test statistic $\alpha'$, because the proportion of male coauthors would be elevated equally for men and women (and so $\alpha'$ would not change). You might ask, what if the super-author has preference for one gender of coauthors -- wouldn't that skew the results somehow? But then, $\alpha'$ would be working as intended: it would correctly record that homophilous (or heterophilous) co-authorships were more common than expected under random assortment by gender. This highlights a feature of $\alpha'$ noted earlier by the reviewer: it is a population-level measure, so we cannot make statements about individuals, e.g. we don't know if everyone is a little bit homophilous, or if there are a few prolific and very homophilous people (though the former is probably closest to the truth, since super-prolific authors with atypical levels of homophily are presumably the exception not the rule).**

@@

#### Discussion
- line 173 : "We found evidence that researchers preferentially publish with same-gendered coauthors" I do not think that this conclusion is adequate given your data. You found evidence that there is a correlation between the genders of authors, but your data does not provide evidence regarding the behavior of individual authors.
Even if there were also such a correlation at the individual level, your data does not allow you to identify the cause of this correlation -- a preference (implying some choice) or a by-product of other constraints.

**We originally used the word 'preference' a bit incautiously, and now use terms like 'assortment' that are more agnostic about the mechanism. We do think individual-level preferences are likely to be partly responsible, however, and so we discuss preference in a few places, while spelling out that it is one possible interpretation for the results.**

### 2018-10-29 manuscript/Manuscript_Claire.Rmd
Resubmission to PLOS ONE

---
title: Researchers preferentially collaborate with same-gendered colleagues across
  the life sciences
author:
- Luke Holman* and Claire Morandin\textsuperscript{§}
- '*\textit{luke.holman@unimelb.edu.au}'
- \textsuperscript{§}\textit{claire.morandin@helsinki.fi} \vspace{5mm}
bibliography: references.bib
output:
  pdf_document: 
    keep_tex: true
  html_document: default
  word_document: default
fontsize: 12pt
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Gender and coauthorship} \usepackage[noabbrev,capitalise]{cleveref} \usepackage{titlefoot}
  \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno} \linenumbers
csl: plos.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious biases, and often feel excluded from formal and informal opportunities for research collaboration. In addition to ensuring fairness and helping to redress gender imbalance in the academic workforce, increasing women's access to collaboration could help scientific progress by drawing on more of the available human capital. Here, we test whether researchers preferentially collaborate with same-gendered colleagues, using more stringent methods and a larger dataset than in past work. Our results reaffirm that researchers preferentially co-publish with colleagues of the same gender, and show that this 'gender homophily' is slightly stronger today than it was 10 years ago. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is stronger in fields where women are in the minority. Interestingly, journals with a high impact factor for their discipline tended to have comparatively low homophily, as predicted if mixed-gender teams produce better research. We discuss some potential causes of gender homophily in science.
  \vspace{5mm} \par\noindent
  \textbf{Keywords:} Gender bias, Homophily, Research collaboration, Text mining, Women in STEM.
---

@@

# Load some little files with numbers to be inserted into the MS
# sample.size <- read.csv("sample.size.table.csv", stringsAsFactors = FALSE)
# sample.size[,2] <- prettyNum(sample.size[,2], big.mark = ",")
# sample.size[,3] <- prettyNum(sample.size[,3], big.mark = ",")
# names(sample.size)[2:3] <- c("n2005", "n2015")

>>>>

---
title: Researchers collaborate with same-gendered colleagues more often than expected across the life sciences
author:
- Luke Holman* and Claire Morandin\textsuperscript{§}
- '*\textit{luke.holman@unimelb.edu.au}'
- \textsuperscript{§}\textit{claire.morandin@helsinki.fi} \vspace{5mm}
bibliography: references.bib
output:
  pdf_document: 
    keep_tex: true
  html_document: default
  word_document: default
fontsize: 12pt
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Gender and coauthorship} \usepackage[noabbrev,capitalise]{cleveref} \usepackage{titlefoot}
  \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno} \linenumbers
csl: plos.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious biases, and often feel excluded from formal and informal opportunities for research collaboration. In addition to ensuring fairness and helping to redress gender imbalance in the academic workforce, increasing women's access to collaboration could help scientific progress by drawing on more of the available human capital. Here, we test whether researchers tend to collaborate with same-gendered colleagues, using more stringent methods and a larger dataset than in past work. Our results reaffirm that researchers co-publish with colleagues of the same gender more often than expected by chance, and show that this 'gender homophily' is slightly stronger today than it was 10 years ago. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is stronger in fields where women are in the minority. Interestingly, journals with a high impact factor for their discipline tended to have comparatively low homophily, as predicted if mixed-gender teams produce better research. We discuss some potential causes of gender homophily in science.
  \vspace{5mm} \par\noindent
  \textbf{Keywords:} Gender bias, Homophily, Research collaboration, Text mining, Women in STEM.
---

@@
